plantecophys - an R package for analysing and modelling leaf gas exchange data

### 2015-09-10 manuscript.Rmd
Minor fixes

\
\* *Corresponding author*
Email: r.duursma@westernsydney.edu.au

*Short title*: The plantecophys R package

@@

Here I present the R package 'plantecophys', a toolkit to analyse and model leaf gas exchange data. Measurements of leaf photosynthesis and transpiration are routinely collected with portable gas exchange instruments, and analyzed with a few key models. These models include the Farquhar-von Caemmerer-Berry (FvCB) model of leaf photosynthesis, the Ball-Berry models of stomatal conductance, and the coupled leaf gas exchange model which combines the supply and demand functions for CO~2~ in the leaf. The 'plantecophys' R package includes functions  for fitting these models to measurements, as well as simulating from the fitted models to aid in interpreting experimental data. Here I describe the functionality and implementation of the new package, and give some examples of its use.  I briefly describe functions for fitting the FvCB model of photosynthesis to measurements of photosynthesis-CO~2~ response curves ('A-C~i~ curves'), fitting Ball-Berry type models, modelling C3 photosynthesis with the coupled photosynthesis-stomatal conductance model, modelling C4 photosynthesis, numerical solution of optimal stomatal behaviour, and energy balance calculations using the Penman-Monteith equation. This open-source package makes technically challenging calculations easily accessible for many users and is freely available on CRAN.

@@

\
We do not repeat a detailed description of the FvCB model here, as it has been described many times [e.g. @bernacchi2001; @medlyn2002]. But generally it is of the form,

@@

\
In the practical application of the FVcB model, it is often assumed that C~c~ is equal to the intercellular [CO~2~] concentration (C~i~), which is useful because C~i~ can be estimated from leaf gas exchange data [@sharkey1982], and this avoids the need to estimate the conductance from intercellular airspaces to the chloroplast (mesophyll conductance, g~m~). In the remainder of this article I will use C~c~ as the driver of photosynthesis, but point out that this can be replaced by C~i~ if the user does not have an estimate of g~m~. When g~m~ is known, C~c~ is calculated from C~i~ with Eq. 2.

@@

\
Fitting the FVcB model to data requires some finesse because net photosynthesis is modelled as a minimum function of two non-linear equations that is sometimes difficult to fit. Moreover, sample sizes collected are often small due to time constraints in the field. A widely used published method requires the user to specify the transition of V~cmax~ to J~max~ limitation [@sharkey2007], a process that is both arbitrary and prevents batch analysis. Another method [@leafweb] requires online submission of data and fits the model without much control or knowledge of the fitting process (following [@gu2010]). Undoubtedly many more implementations of the fitting process have been developed over the years, but few of these are made publicly available (but see available online tools [@dekauwefitfarquhar; @landfluxtools]). What is missing is an open-source tool that can be used for reproducible and transparent analysis of A-C~i~ curves.

@@

\
Despite the widespread use of the FvCB model and the coupled leaf gas exchange model, tools to analyze data and perform simulations are scattered and subject to little standardization. Fitting the FvCB model to CO~2~ response curves is a standard procedure but different methods can yield different parameter values, making comparisons difficult. The coupled leaf gas exchange model is not straightforward to implement, and I do not know of any standalone open-source implementations. I here describe the plantecophys package, implemented in the R language [@rfoundation]. The code is freely available (without restrictions), and managed with a version control system. The package is the result of our work on leaf and canopy modelling of photosynthesis and stomatal conductance [@medlyn2011; @barton2012; @peltoniemi2012; @medlyn2013; @duursma2013; @duursma2014; @lin2015], with many additions based on user requests.

@@

The straightforward `fitBB` function provides an interface to non-linear or linear regression to fit one of three stomatal conductance models [@ballberry1987; @leuning1995; @medlyn2011]. This yields estimates of g~1~ and (optionally ) g~0~, which are necessary inputs to the coupled leaf gas exchange model. Note that the user must provide stomatal conductance to H~2~O (not CO~2~) as input to the fitting process, which is the standard output of portable gas exchange instruments. This function is demonstrated in the example application further below.

@@

\
The `Photosyn` function assumes that the boundary layer conductance (g~bl~) is high compared to g~s~, so that T~leaf~ is close to T~air~. As an alternative, the `PhotosynEB` function calculates T~leaf~ from the leaf energy balance. Transpiration is calculated with the Penman-Monteith equation, which accounts for boundary layer effects. The details of `PhotosynEB` are not described here (we refer to the documentation), because it is very similar to other implementations [@wang1998; @buckley2014].

@@

```
# Fit A-Ci curves. In this case, each separate curve is indexed by a column names 'Curve',
# and the data were already read into a dataframe (tumh)
acifits <- fitacis(tumh, "Curve")

@@

# Fit Medlyn et al.'s (2011) version of the Ball-Berry model
# Data are already read into a dataframe (tumspot), and have standard names (or they can be set).
gfit <- fitBB(tumspot, gsmodel="BBOpti")

@@

We need an open source set of tools to analyze leaf gas exchange data, as these data form a cornerstone of plant physiological ecology. At the moment there are no publicly available tools to fit A-C~i~ curves that can be used as part of a reproducable workflow. The plantecophys R package is implemented in widely used language for data analysis. The package includes a useful set of tools to perform standard, and more advanced, analyses of leaf gas exchange data. The open source framework combined with version control allows further development of the code.

>>>>

\
\* *Corresponding author*
\
Email: r.duursma@westernsydney.edu.au

\
*Short title*: The plantecophys R package

@@

Here I present the R package 'plantecophys', a toolkit to analyse and model leaf gas exchange data. Measurements of leaf photosynthesis and transpiration are routinely collected with portable gas exchange instruments, and analysed with a few key models. These models include the Farquhar-von Caemmerer-Berry (FvCB) model of leaf photosynthesis, the Ball-Berry models of stomatal conductance, and the coupled leaf gas exchange model which combines the supply and demand functions for CO~2~ in the leaf. The 'plantecophys' R package includes functions  for fitting these models to measurements, as well as simulating from the fitted models to aid in interpreting experimental data. Here I describe the functionality and implementation of the new package, and give some examples of its use.  I briefly describe functions for fitting the FvCB model of photosynthesis to measurements of photosynthesis-CO~2~ response curves ('A-C~i~ curves'), fitting Ball-Berry type models, modelling C3 photosynthesis with the coupled photosynthesis-stomatal conductance model, modelling C4 photosynthesis, numerical solution of optimal stomatal behaviour, and energy balance calculations using the Penman-Monteith equation. This open-source package makes technically challenging calculations easily accessible for many users and is freely available on CRAN.

@@

\
I do not repeat a detailed description of the FvCB model here, as it has been described many times [e.g. @bernacchi2001; @medlyn2002]. But generally it is of the form,

@@

\
In the practical application of the FvCB model, it is often assumed that C~c~ is equal to the intercellular [CO~2~] concentration (C~i~), which is useful because C~i~ can be estimated from leaf gas exchange data [@sharkey1982], and it avoids the need to estimate the conductance from intercellular airspaces to the chloroplast (mesophyll conductance, g~m~). In the remainder of this article I will use C~c~ as the driver of photosynthesis, but point out that this can be replaced by C~i~ if the user does not have an estimate of g~m~. When g~m~ is known, C~c~ is calculated from C~i~ with Eq. 2.

@@

\
Fitting the FvCB model to data requires some finesse because net photosynthesis is modelled as a minimum function of two non-linear equations that is sometimes difficult to fit. Moreover, sample sizes collected are often small due to time constraints in the field. A widely used published method requires the user to specify the transition of V~cmax~ to J~max~ limitation [@sharkey2007], a process that is both arbitrary and prevents batch analysis. Another method [@leafweb] requires online submission of data and fits the model without much control or knowledge of the fitting process (following [@gu2010]), and does not report standard errors of the estimated parameters. Undoubtedly many more implementations of the fitting process have been developed over the years, but few of these are made publicly available (but see available online tools [@dekauwefitfarquhar; @landfluxtools]). What is missing is an open-source tool that can be used for reproducible and transparent analysis of A-C~i~ curves.

@@

\
Despite the widespread use of the FvCB model and the coupled leaf gas exchange model, tools to analyse data and perform simulations are scattered and subject to little standardization. Fitting the FvCB model to CO~2~ response curves is a standard procedure but different methods can yield different parameter values, making comparisons difficult. The coupled leaf gas exchange model is not straightforward to implement, and I do not know of any standalone open-source implementations. I here describe the plantecophys package, implemented in the R language [@rfoundation]. The code is freely available (without restrictions), and managed with a version control system. The package is the result of our work on leaf and canopy modelling of photosynthesis and stomatal conductance [@medlyn2011; @barton2012; @peltoniemi2012; @medlyn2013; @duursma2013; @duursma2014; @lin2015], with many additions based on user requests.

@@

The straightforward `fitBB` function provides an interface to non-linear or linear regression to fit one of three stomatal conductance models [@ballberry1987; @leuning1995; @medlyn2011]. This yields estimates of g~1~ and (optionally) g~0~, which are necessary inputs to the coupled leaf gas exchange model. Note that the user must provide stomatal conductance to H~2~O (not CO~2~) as input to the fitting process, which is the standard output of portable gas exchange instruments. This function is demonstrated in the example application further below.

@@

\
The `Photosyn` function assumes that the boundary layer conductance (g~bl~) is high compared to g~s~, so that T~leaf~ is close to T~air~. As an alternative, the `PhotosynEB` function calculates T~leaf~ from the leaf energy balance. Transpiration is calculated with the Penman-Monteith equation, which accounts for boundary layer effects. The details of `PhotosynEB` are not described here (see the built in help file for more information), because it is very similar to other implementations [@wang1998; @buckley2014].

@@

```
# Fit A-Ci curves. 
# In this case, each separate curve is indexed by a column named 'Curve',
# and the data were already read into a dataframe (tumh)
acifits <- fitacis(tumh, "Curve")

@@

# Fit Medlyn et al.'s (2011) version of the Ball-Berry model
# Data are already read into a dataframe (tumspot), 
# and have standard names (or they can be set).
gfit <- fitBB(tumspot, gsmodel="BBOpti")

@@

We need an open source set of tools to analyse leaf gas exchange data, as these data form a cornerstone of plant physiological ecology. At the moment there are no publicly available tools to fit A-C~i~ curves that can be used as part of a reproducable workflow. The plantecophys R package is implemented in widely used language for data analysis. The package includes a useful set of tools to perform standard, and more advanced, analyses of leaf gas exchange data. The open source framework combined with version control allows further development of the code.
